---
title: "Simulation"
output:
  pdf_document: default
  html_notebook: default
---

# Three girls

## Simulating one family with random numbers

A single random number:

```{r}
runif(1)
```

Look for the R help on `runif`.  `runif` makes Random UNIForm numbers; random numbers from the [uniform distribution](https://en.wikipedia.org/wiki/Uniform_distribution_(continuous)).  Uniform means that all numbers between 0 and 1 are equally likely.

We start by simulating a coin toss.

```{r}
runif(1) < 0.5
```

`runif` needs a single argument: the number of random numbers we want.

Here we get a vector of four random numbers:

```{r}
runif(4)
```

We use our *conditional expression* skills to simulate four coin tosses:

```{r}
runif(4) < 0.5
```

Do the same thing again, but keep the result:

```{r}
coin_tosses <- runif(4) < 0.5
coin_tosses
```

We would like to know the number of `TRUE` values.

It turns out that R, like many other languages, thinks of `TRUE` as being equal to 1, and `FALSE` as equal to 0:

```{r}
TRUE + TRUE
```

```{r}
FALSE + FALSE
```

```{r}
TRUE + FALSE
```

This means that we can use the `sum` function on our vector of coin tosses:

```{r}
coin_tosses
```

```{r}
sum(coin_tosses)
```

Make another vector to check again that we get the right answer from `sum`:

```{r}
coin_tosses_2 <- runif(4) < 0.5
coin_tosses_2
```

```{r}
sum(coin_tosses_2)
```

## Simulating five families

So far we have simulated four coin tosses.  Each `TRUE` value represents a girl.  This represents one family.

What if we want to do this five times, to simulate five families.  We could do this a very long and boring way.

First I make a vector of all zero, to store the counts:

```{r}
counts <- numeric(5)
counts
```

Then I trudge through making five families one after the other:

```{r}
coin_tosses_1 <- runif(4) < 0.5
counts[1] <- sum(coin_tosses_1)
coin_tosses_2 <- runif(4) < 0.5
counts[2] <- sum(coin_tosses_2)
coin_tosses_3 <- runif(4) < 0.5
counts[3] <- sum(coin_tosses_3)
coin_tosses_4 <- runif(4) < 0.5
counts[4] <- sum(coin_tosses_4)
coin_tosses_5 <- runif(4) < 0.5
counts[5] <- sum(coin_tosses_5)
```

I have one count for each simulated family:

```{r}
counts
```

How many of these are equal to three?  I can use my vector skills again:

```{r}
counts_equal_3 <- counts == 3
counts_equal_3
```

Now I can count how many are equal to three:

```{r}
sum(counts_equal_3)
```

In fact I can do this all in one go, by putting the `counts == 3` expression inside the parentheses:

```{r}
sum(counts == 3)
```

R first finds the stuff between the parentheses.  It *evaluates* the stuff between parentheses, to get the vector of `TRUE` and `FALSE`.  It passes the result to `sum`.

The big chunk simulating four families looks very repetitive.  What happens if I need to do 20 families?  Or 1000?  I can't do it like that.

We need a way to do the same thing multiple times.  In our case, we need a way to make a family, and count the number of girls, many times.  To do this, we will use a `for` loop.

## print

To explain `for` loops, we will have use of another function in R: `print`.

`print` is a simple function that takes an argument, and displays that argument in the notebook or console.

```{r}
print(10)
```

```{r}
print('Hello')
```

You will soon see why we need this, instead of our usual:

```{r}
10
```

or

```{r}
'Hello'
```

## Repeating things many times with `for` loops

A `for` loop allows us to repeat the same thing many times.

Here is an example:

```{r}
for (index in 1:5) {
  print(index)
}
```

Read this as "For every value 1 through 5 - put the value into the variable
`index`, then print the value of `index`".

The `for` loop starts like this: `for (variable in vector)`.  `variable` can be any variable name:

```{r}
for (funky_chicken in 1:5) {
  print(funky_chicken)
}
```

`vector` can be any vector of values:

```{r}
my_vector <- c(3, 1, 4, 1)
my_vector
```

```{r}
for (funky_chicken in my_vector) {
  print(funky_chicken)
}
```

Read `for (funky_chicken in my_vector)` as "For each value in 'my_vector', put the value into variable `funky_chicken` and then do ... " something.

The "something" is the stuff between the curly braces `{}`.  This is called the *loop body*.  It's the stuff that happens once for each value in `my_vector`.  Consider this `for` loop again:

```{r}
for (funky_chicken in my_vector) {
  print(funky_chicken)
}
```

The *body* of the loop is stuff between the curly braces. In this case it is just the line `print(funky_chicken)`.

Put another way, the `for` loop above does exactly the same thing as:

```{r}
funky_chicken <- my_vector[1]
# The stuff in the body of the for loop
print(funky_chicken)
funky_chicken <- my_vector[2]
# The stuff in the body of the for loop
print(funky_chicken)
funky_chicken <- my_vector[3]
# The stuff in the body of the for loop
print(funky_chicken)
funky_chicken <- my_vector[4]
# The stuff in the body of the for loop
print(funky_chicken)
funky_chicken <- my_vector[5]
# The stuff in the body of the for loop
print(funky_chicken)
```

You might agree that the `for` loop version looks a bit neater.  Even if you don't, you can imagine that we can't do the version above if we have too many elements in our vector.

With that introduction, we return to our ugly version of simulating the five families:

```{r}
# Make a vector 5 long with all zeros
counts <- numeric(5)
# Make five families, and count the number of girls
coin_tosses_1 <- runif(4) < 0.5
counts[1] <- sum(coin_tosses_1)
coin_tosses_2 <- runif(4) < 0.5
counts[2] <- sum(coin_tosses_2)
coin_tosses_3 <- runif(4) < 0.5
counts[3] <- sum(coin_tosses_3)
coin_tosses_4 <- runif(4) < 0.5
counts[4] <- sum(coin_tosses_4)
coin_tosses_5 <- runif(4) < 0.5
counts[5] <- sum(coin_tosses_5)
# Show the counts
counts
```

We rewrite with a `for` loop:

```{r}
counts <- numeric(5)
for (i in 1:5) {
  coin_tosses <- runif(4) < 0.5
  counts[i] <- sum(coin_tosses)
}
counts
```

The trick here is that we can do this as many times as we like.  Let's start with 50 times:

```{r}
counts <- numeric(50)
for (i in 1:50) {
  coin_tosses <- runif(4) < 0.5
  counts[i] <- sum(coin_tosses)
}
counts
```

Now 10000 times:

```{r}
n_repeats <- 10000
counts <- numeric(n_repeats)
for (i in 1:n_repeats) {
  coin_tosses <- runif(4) < 0.5
  counts[i] <- sum(coin_tosses)
}
# Show the first 10
head(counts, 10)
```

How many times do we see a count of exactly 3?

```{r}
number_of_threes <- sum(counts == 3)
number_of_threes
```

What proportion is this?

```{r}
number_of_threes / n_repeats
```

## Summary

If you want to repeat something `n` times:

```{r}
# For example
n <- 10
```

Then:

```{r}
for (i in 1:n) {
  # The thing you want to repeat
}
```
